---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1:** The name is in single quotation marks instead of double quotation marks.

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** There are 10,752 data points and 22 columns, giving us date/time/location and technical statistics about earthquakes in 2026.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |>
  arrange(desc(depth))

```

**A3:**The deepest two earthquakes were at 750 depth - one had an N/A magnitude and the other had a 3.6 magnitude

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes |>
  filter(mag > 6)
```

**A4:** 12 earthquakes fit that criteria.

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6) |>
  filter(depth < 20)

```

**A5:** 10 earthquakes have a magnitude above 6 and depth below 20.

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6|depth < 20)
```

**A6:**8,544 earthquakes have *either* a magnitude above 6 or depth below 20.

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes |>
  filter(str_detect(place, "Alaska"))
```

**A7:** You can look at the column "place" and use str_detect to identify places with the word "Alaska" in the name. There are 3,578 Alaskan earthquakes in the dataset.

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

**A8:** The best way to find all the versions of types is to use group_by and the row-counting n() function. Earthquake is by far the most common, with 10,580, and quarry blast is second with 105.

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(average = mean(depth)) |>
  arrange(desc(average))
```

**A9:** The average earthquake depth is 22.785. It's somewhat unusual to me that ice quake's average depth is *exactly* zero — I'm sure that ice can exist above or below whatever we are baselining depth at.

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  group_by(hour) |>
  summarise(total = n()) |>
  arrange(desc(total))
  
```

**A10:** Use the mutate function as well as the lubridate "hour" function to create a new column with the hour the earthquake took place, counting from 0 to 23.zA

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
earthquakes |>
  mutate(over_three = case_when(
    mag > 3 ~ "Yes",
    mag <= 3 ~ "No",
    .default = NA
  )) |>
  arrange(desc(over_three))
    
```

**A11:** Use mutate() to create a new column indicating whether an earthquake's magnitude is above three or not, in plain terms.

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  mutate(over_three = case_when(
    mag > 3 ~ "Yes",
    mag <= 3 ~ "No",
    .default = NA
  )) |>
  filter(str_detect(over_three, "Yes")) |>
  group_by(hour) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

**A12**: *need to ask about this in class because I have zero clue where to start*

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?

```{r}
#Q - do more earthquakes happen above or below the Equator?
earthquakes |>
  mutate(above_below = case_when(
    latitude > 0 ~ "above",
    latitude < 0 ~ "below",
    .default = NA)) |>
  group_by(above_below) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

```{r}
#Q - Are earthquakes stronger above or below the equator?
earthquakes |>
  mutate(above_below = case_when(
    latitude > 0 ~ "above",
    latitude < 0 ~ "below",
    .default = NA)) |>
  group_by(above_below) |>
  summarise(mean_mag = mean(mag)) |>
  arrange(desc(mean_mag))
```

```{r}
#Is the magnitude error for reviewed earthquakes smaller than error for automatically tracked earthquakes?
earthquakes |>
  group_by(status) |>
  summarise(mean = mean(magError)) |>
  arrange(desc(mean))
```
