---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

glimpse(lab_01)
## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1** The sole CMNS class has 440 seats, which is the highest average. Given CMNS college has a ton of departments within it, I wonder if this is a freshman intro class or something. FIRE and the engineering education program being 2 and 3 gives credence to freshman intro class theory. Air science classes being massive surprises me as well.

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats),
  ) |>
  arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2**COMM107 has been taught 1400 times in the dataset.

```{r}
umd_courses |>
  group_by(id) |>
  summarise(
    total_classes = sum(sections),
  ) |>
  arrange(desc(total_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3**Chemistry professor Monique Koppel has taught a UMD-record 611 courses.


**there is definitely a better way to do this and I will ask about it in class tomorrow - can't figure out how to split up instructors that are grouped together**
```{r}
umd_courses |>
  group_by(instructors, department) |>
  summarise(
    total_classes_taught = sum(sections),
  ) |>
  arrange(desc(total_classes_taught))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4** The data needs to be made into a table; or at minimum an extensive spreadsheet with some pivot tables. There's entirely too much data to sift through in the format it's presented right now. I also couldn't figure out how to split up professors who are condensed into one line. But i'm 99% confident that Koppel is the right answer.

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5** Pre-lab Q: Which departments have the fewest seats relative to total # of classes?
        Excluding the National Institute of Aeronautics in North Carolina(???), music performances has the fewest seats per class.


```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    seats_per_class = sum(seats) / n()
  ) |>
  arrange(seats_per_class)
```
